---
title: "OC17_MAGs_refined_annotation"
output: html_document
date: "2024-06-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(data.table)
library(tidyr)
library(dplyr)
library(readr)
library(purrr)
library(ggplot2)
library(patchwork)
library(ggthemes)
library(RColorBrewer)
library(ggh4x)
```

```{r}
rootdir = "/scratch/users/rsalcedo/OC17wc_deepauto/genome_based/"
genomedir = paste(rootdir, "anvio/refined_bins/", sep = "")
seqdir = "/scratch/users/rsalcedo/OC17wc_sequences/"
sample_list = c("3000m1000m", "3000m150m", "3000m2000m", "3000m3000m", "3000m500m", "3000m50m", "3500m1000m", "3500m150m", "3500m2000m", "3500m3000m", "3500m500m", "3500m50m", "4000m1000m", "4000m150m", "4000m2000m", "4000m3000m", "4000m500m", "4000m50m", "4500m150m", "4500m2000m", "4500m3000m", "4500m4000m", "4500m500m", "4500m50m", "500m150m", "500m500m", "500m50m", "4500m1000m")
bin_list = gsub("_contigs.fa", "", list.files(path = genomedir))
```

CheckM
```{r}
#make sure to activate base_metagenomics conda environment
sink(paste(rootdir, "refined_checkM.sh", sep = ""))

cat("sbatch -J RcheckM -c 32 --mem-per-cpu 16G -p serc -t 160:00:00 --wrap 'checkm lineage_wf -t 32 --pplacer_threads 32 -x fa --tab_table ", genomedir, " ", rootdir, "refined_checkM/'", "\n", sep = "")

sink()

file.show(paste(rootdir, "refined_checkM.sh", sep = ""))
```

Parsing checkM
```{r}
sink(paste(rootdir, "refined_checkM_qa.sh", sep = ""))
cat("sbatch -J Rqa -c 20 --mem-per-cpu 8G -p serc -t 160:00:00 --wrap 'checkm qa -t 20 -o 2 -f ", rootdir, "refined_checkM/refined_output_table.txt --tab_table ", rootdir, "refined_checkM/lineage.ms ", rootdir, "refined_checkM/'", "\n", sep = "")

sink()

file.show(paste(rootdir, "refined_checkM_qa.sh", sep = ""))
```

prodigal
```{r}
sink(paste(rootdir, "refined-prodigal.sh", sep = ""))
for (i in bin_list) {
  cat("sbatch -J ", i, " -c 20 --mem-per-cpu 8G -p serc -t 160:00:00 --wrap 'prodigal -i ", genomedir, i, "_contigs.fa -d ", rootdir, "refined_prodigal/fna/", i, "_prodigalORFs.fna -a ", rootdir, "refined_prodigal/faa/", i, "_prodigalORFs.faa -o ", rootdir, "refined_prodigal/log/", i, "_prodigal.log'\n", sep = "")
  }
sink()

file.show(paste(rootdir, "refined-prodigal.sh", sep = ""))
```

kofamscan
```{r}
#remember to activate kofamscan conda environment

kofamscan_path = "/home/groups/dekas/rebecca/kofamscan/kofam_scan-1.3.0/exec_annotation"
ko_list = "/home/groups/dekas/rebecca/kofamscan/ko_list"
ko_profiles = "/home/groups/dekas/rebecca/kofamscan/profiles"

sink(paste(rootdir, "refined_kofamscan.sh", sep = ""))
cat("#!/bin/bash\n", "#SBATCH --job-name=koscan\n", "#SBATCH -p serc\n", "#SBATCH --time=168:00:00\n", "#SBATCH --ntasks=1\n", "#SBATCH --cpus-per-task=32\n", "#SBATCH --mem-per-cpu=8G\n\n", "for i in $(cat ", paste(rootdir, "refined_bin_list.txt", sep = ""), ")\n", "do\n", kofamscan_path, " -o ", rootdir, "refined_kofamscan/txt/${i}_KEGGIDs.txt -k ",  ko_list, " -p ", ko_profiles, " --cpu 32 --tmp-dir ", rootdir, "tmpdirs/${i}_tmpdir/ ", rootdir, "refined_prodigal/faa/${i}_prodigalORFs.faa\n", "done\n", sep = "")
sink()

file.show(paste(rootdir, "refined_kofamscan.sh", sep = ""))
```

kofamparse
```{r}
#remember to activate base_metagenomics conda environment and do ml ruby

kofamparse_path = "/home/groups/dekas/software/kofamparse/kofamparse"
sink(paste(rootdir, "refined_kofamparse.sh", sep = ""))
  cat("#!/bin/bash\n", "#SBATCH --job-name=koparse\n", "#SBATCH -p serc\n", "#SBATCH --time=168:00:00\n", "#SBATCH --ntasks=1\n", "#SBATCH --cpus-per-task=32\n", "#SBATCH --mem-per-cpu=8G\n\n", "for i in $(cat ", paste(rootdir, "refined_bin_list.txt", sep = ""), ")\n", "do\n", kofamparse_path, " ", rootdir, "refined_analysis/refined_kofamscan/txt/${i}_KEGGIDs.txt ", rootdir, "refined_analysis/refined_kofamscan/csv/${i}_KEGGIDs.csv\n", "done\n", sep = "")
sink()

file.show(paste(rootdir, "refined_kofamparse.sh", sep = ""))

```

coverM
map to whole drep set but replace old mags with refined versions before running coverM
```{r}
sink(paste(rootdir,"refined_analysis/merged_coverM.sh", sep = ""))
for (i in sample_list) {
  cat("sbatch -J ", i, "_coverM -c 20 --mem-per-cpu 8G -p serc -t 160:00:00 --wrap 'coverm genome --coupled ", seqdir, "trimmed_reads/", i, "_R1_trimmed.fastq.gz ", seqdir, "trimmed_reads/", i, "_R2_trimmed.fastq.gz --genome-fasta-directory ",  rootdir, "merged_genome_set/ -x fa -t 20 --min-read-percent-identity 95 --output-format sparse --min-covered-fraction 0 -m count mean covered_fraction length -o ", rootdir, "refined_analysis/merged_coverM/", i, "_coverM.tsv'\n", sep = "")
  }
sink()

file.show(paste(rootdir,"refined_analysis/merged_coverM.sh", sep = ""))
```

parsing coverM and merging 
```{r message=FALSE, warning=FALSE}
for(i in sample_list) {                                           
  coverM = read_tsv(paste(rootdir, "refined_analysis/merged_coverM/", i, "_coverM.tsv", sep = ""), col_names = TRUE)
  coverM$Sample = gsub("_R1_trimmed.fastq.gz", "", coverM$Sample)
  coverM = rename(coverM, genome = Genome)
  coverM$genome = gsub("_contigs", "", coverM$genome)
  coverM = rename(coverM, covered_fraction = "Covered Fraction")
  assign(paste("coverM_", i, sep = ""), coverM)
}


df_list = mget(ls(pattern = "coverM_"))
coverM = rbindlist(df_list, use.names = TRUE)

write_csv(coverM, paste(rootdir, "refined_analysis/refined_coverM_all.csv", sep = ""))
```

kofamscan thresholding
```{r message=FALSE, warning=FALSE}
KO_thresholds = read_csv(paste(rootdir, "bitscore_thresholds.csv", sep = ""))

#define list of KO's im interested in
KOs = list("K00194", "K00197", "K00198", "K00855", "K01601", "K01602", "K08691", "K09709", "K14466", "K14467", "K14468", "K14469", "K14470", "K15016", "K15017", "K15023", "K15039", "K15052", "K15230", "K15231", "K15232", "K15233", "K15234", "K18861", "K14138")

for(i in bin_list) {                                           
  kegg = read_csv(paste(rootdir, "refined_kofamscan/csv/", i, "_KEGGIDs.csv", sep = ""), col_names = TRUE)
  kegg$score = as.integer(kegg$score)
  kegg$threshold = as.integer(kegg$threshold)
  kegg = rename(kegg, gene = "Gene name") 
  kegg = rename(kegg, KO = "KO number") 
  kegg = rename(kegg, evalue = "e-value") 
  kegg_besthit = kegg %>% 
    group_by(gene) %>% 
    slice_max(score)
  kegg_besthit = subset(kegg_besthit, kegg_besthit$evalue < 1e-6)
  kegg_besthit = subset(kegg_besthit, kegg_besthit$score >= threshold)
  
  auto_kegg = subset(kegg_besthit, (KO %in% KOs))
  for (x in KOs) {
    threshold = with(KO_thresholds, new[KO == x])
    assign(paste("bitscore_", x, sep = ""), threshold)
    pass <- auto_kegg[which(auto_kegg$KO == x & auto_kegg$score >= threshold), ]
    assign(paste("bitscore_pass_", x, sep = ""), pass)
  }
  auto_filtered <- rbind(bitscore_pass_K00194, bitscore_pass_K00197, bitscore_pass_K00198, bitscore_pass_K00855, bitscore_pass_K01601, bitscore_pass_K01602, bitscore_pass_K08691, bitscore_pass_K09709, bitscore_pass_K14466, bitscore_pass_K14467, bitscore_pass_K14468, bitscore_pass_K14469, bitscore_pass_K14470, bitscore_pass_K15016, bitscore_pass_K15017, bitscore_pass_K15023, bitscore_pass_K15039, bitscore_pass_K15052, bitscore_pass_K15230, bitscore_pass_K15231, bitscore_pass_K15232, bitscore_pass_K15233, bitscore_pass_K15234, bitscore_pass_K18861, bitscore_pass_K14138)
  auto_filtered  = auto_filtered  %>%
    mutate(pathway = case_when(
      endsWith(KO, "00855") ~ "CBB",
      endsWith(KO, "01601") ~ "CBB",
      endsWith(KO, "01602") ~ "CBB",
      endsWith(KO, "15230") ~ "rTCA", 
      endsWith(KO, "15231") ~ "rTCA",
      endsWith(KO, "15232") ~ "rTCA",
      endsWith(KO, "15233") ~ "rTCA",
      endsWith(KO, "15234") ~ "rTCA",
      endsWith(KO, "00198") ~ "WL",
      endsWith(KO, "14138") ~ "WL",
      endsWith(KO, "00197") ~ "WL",
      endsWith(KO, "00194") ~ "WL",
      endsWith(KO, "15023") ~ "WL",
      endsWith(KO, "14468") ~ "3HP Bicycle",
      endsWith(KO, "15052") ~ "3HP Bicycle",
      endsWith(KO, "08691") ~ "3HP Bicycle",
      endsWith(KO, "14469") ~ "3HP Bicycle",
      endsWith(KO, "14470") ~ "3HP Bicycle",
      endsWith(KO, "09709") ~ "3HP Bicycle",
      endsWith(KO, "15039") ~ "3HP/4HB",
      endsWith(KO, "14466") ~ "3HP/4HB",
      endsWith(KO, "15017") ~ "3HP/4HB or DC/4HB",
      endsWith(KO, "15016") ~ "3HP/4HB or DC/4HB",
      endsWith(KO, "18861") ~ "3HP/4HB or DC/4HB",
      endsWith(KO, "14467") ~ "DC/4HB"
  ))
  duprows <- rownames(kegg_besthit) %in% rownames(auto_filtered)
  processed = rbind(auto_filtered, kegg_besthit[!duprows,])
  write.csv(processed, paste(rootdir, "refined_kofamscan/processed/", i, "_processed.csv", sep = ""))
}

```

Figure 4: metabolic potential of autotrophic MAGs
```{r message=FALSE, warning=FALSE}
DRAM_KOs = read_tsv(paste(rootdir, "DRAM_energy_KOs_v5.tsv", sep = ""))
DRAM_pathways = read_tsv(paste(rootdir, "DRAM_energy_pathways_v5.tsv", sep = ""), col_names = TRUE)
checkM_info = read_tsv(paste(rootdir, "refined_analysis/refined_checkM/refined_output_table.txt", sep = ""))
gtdb_info = read_tsv(paste(rootdir, "gtdbtk_summary.tsv", sep = ""))
coverM = read_csv(paste(rootdir, "refined_analysis/refined_coverM_all.csv", sep = ""))
```

read in all the annotations, merge them with DRAM information and save as new df
```{r message=FALSE, warning=FALSE}
filenames = list.files(path = "/scratch/users/rsalcedo/OC17_genomes/refined_analysis/refined_kofamscan/processed/")
setwd("/scratch/users/rsalcedo/OC17_genomes/refined_analysis/refined_kofamscan/processed/")

for (f in filenames){
  genome<-read_csv(f, col_names = TRUE)
  genome = as.data.table(genome$KO)
  genome$present = 1
  genome = genome %>%
    rename(!!f := "present")
  colnames(genome)[2] <-  sub("\\_processed.csv$", "", colnames(genome)[2])
  genome = genome %>%
    rename("gene_id" = "V1")
  genome = subset(genome, !duplicated(subset(genome, select=gene_id)))
  genome = left_join(DRAM_KOs, genome, by = "gene_id")
  assign(paste("genome", f, sep="_"), genome)
}

```

merge each genomes df's together into one giant matrix of gene presence/absence and clean up
```{r}
#get all data frames in list
df_list = mget(ls(pattern = "genome_R"))
#clean up the names so its just the bin ID
names(df_list) <- sub("\\_processed.csv$", "", names(df_list))
#merge all the dataframes together
a = do.call("cbind", df_list)

#clean up df by renaming some files and dropping duplicate columns
a = a %>%
    rename("KO" = "genome_R3000m1000m.141.gene_id")
ab <- a %>% select(-contains(".gene_id"))

#reorder the KO columns so theyre in the same order for when you do cbind since that doesnt look for matches. had to do it this way since merge wasnt working (gave too many rows)
DRAM_pathways = DRAM_pathways[order(DRAM_pathways$gene_id,decreasing=TRUE),]
ab = ab[order(ab$KO,decreasing=TRUE),]
ab = cbind(ab, DRAM_pathways)

#dropping duplicate rows, renaming column headers, and replacing NAs with zeros
ab <- ab %>% select(-contains("gene_id"))
ab[is.na(ab)] <- 0
colnames(ab) = gsub("[.]R\\d+m\\d+m*[.]\\d+", "", colnames(ab))
```

manipulate presence/absence dataframe above to long form so I can add in the taxonomic identity and subset based on checkM quality metrics
```{r}
#now turn ab dataframe into long form so we can add tax and plot as heatmap
ab_long = ab %>% pivot_longer(cols = starts_with("genome_"), names_to = "genome", values_to = "present")
#remove the genome_ part so that it matches the data in the drep_tax df that I need to merge it with
ab_long$genome = sub("genome_", "", ab_long$genome)

#modify gtdb bin names to match new refined bin names
gtdb_info = gtdb_info %>%
    rename("genome" = "user_genome")

gtdb_info = select(gtdb_info, "genome", "classification")

gtdb_info$genome = gsub("A", "R",gtdb_info$genome)

#merge the dataframes
ab_tax = left_join(ab_long, gtdb_info, by = "genome")

#clean up taxonomy names for plotting
ab_tax$phylum = sub("\\;c.*", "", ab_tax$classification)
ab_tax$class = sub("\\;o.*", "", ab_tax$classification)

#add in checkM info
checkM_info = checkM_info %>%
    rename("genome" = "Bin Id")
checkM_info = checkM_info %>%
    rename("size_bp" = "Genome size (bp)")
checkM_info = select(checkM_info, "genome", "Completeness", "Contamination", "size_bp")
checkM_info$genome = gsub("_contigs", "", checkM_info$genome)


ab_tax = left_join(ab_tax, checkM_info, by = "genome")
#drop MAGs with over 5% contamination and under 50% completeness
ab_tax_505 = filter(ab_tax, ab_tax$Completeness >= 50 & ab_tax$Contamination <= 5)

#make a list of genomes with one or more
#start by getting only rows of autotrophy genes
auto = filter(ab_tax_505, ab_tax_505$header == "carbon fixation")
#sum that data by genome to find the total carbon fixation genes in each genome
aa = aggregate(present ~ genome, data=auto, FUN=sum)
#subset the above df to get a list of genomes that have at least two genes involved in carbon fixation
auto_sum = filter(aa, aa$present >= 1)
#now use that list to pull out the mags of interest from the whole dataset
auto_MAGs = subset(ab_tax_505, (genome %in% auto_sum$genome))

length(unique(auto_MAGs$genome))
length(unique(ab_tax_505$genome))

write.csv(auto_MAGs, paste(rootdir, "refined_analysis/refined_auto_MAGs.csv", sep = ""), row.names = FALSE)

```

Table S1: genome, taxonomy, size, GC, carbon fixation pathway present, completeness, contamination, sample generated from, relative abundance, site, depth
```{r}

tableS1 = auto_MAGs
tableS1 = filter(tableS1, tableS1$present == 1)

CBB <- c("K01602", "K00855", "K01601")
HP <- c("K14468", "K15052", "K08691", "K14469", "K14470", "K14449", "K09709")
rTCA = c("K15230", "K15231", "K15232", "K15233", "K15234")
WL = c("K00198", "K14138", "K00197", "K00194", "K15023")
HP_HB = c("K15039", "K14466")
HP_DC_HB = c("K15017", "K15016", "K18861", "K14467")

tableS1$pathway = "X"
tableS1 <- tableS1 %>%
  group_by(genome) %>%
  mutate(pathway = case_when(
    any(KO %in% CBB) & any(KO %in% HP) ~ "3HP bicycle, CBB",
    any(KO %in% CBB) ~ "CBB",
    any(KO %in% HP) ~ "3HP bi-cycle",
    any(KO %in% rTCA) ~ "rTCA",
    any(KO %in% WL) ~ "WL",
    any(KO %in% HP_HB) ~ "3HP/4HB",
    any(KO %in% HP_DC_HB) ~ "3HP/4HB or DC/4HB",
    TRUE ~ NA_character_
  ))

unique(tableS1$pathway)

tableS1$KO = NULL
tableS1$gene_description = NULL
tableS1$header = NULL
tableS1$module = NULL
tableS1$present = NULL

tableS1 = subset(tableS1,!is.na(pathway))

tableS1 = distinct(tableS1, genome, .keep_all = TRUE)

#modify pathway based on taxonomy
tableS1 <- tableS1 %>%
  mutate(
    pathway = case_when(
      grepl("halobacteriota", phylum, ignore.case = TRUE) & pathway == "3HP/4HB or DC/4HB" ~ "DC/4HB",
      grepl("thermoproteota|thermoplasmatota|planctomycetota|myxococcota", phylum, ignore.case = TRUE) & pathway == "3HP/4HB or DC/4HB" ~ "3HP/4HB",
      TRUE ~ pathway
    )
  )

tableS1$depth_generated = tableS1$genome
tableS1$depth_generated = sub("R", "", tableS1$depth_generated)
tableS1$depth_generated = sub("*\\.[0-9]+", "", tableS1$depth_generated)

tableS1  = tableS1  %>%
  mutate(depth_generated = case_when(
    endsWith(depth_generated, "m50m") ~ "50m",
    endsWith(depth_generated, "150m") ~ "150m",
    endsWith(depth_generated, "500m") ~ "500m",
    endsWith(depth_generated, "1000m") ~ "1000m",
    endsWith(depth_generated, "2000m") ~ "2000m",
    endsWith(depth_generated, "3000m") ~ "3000m",
    endsWith(depth_generated, "4000m") ~ "4000m")
  )
    
tableS1$site_generated = tableS1$genome

tableS1$site_generated = sub("R", "", tableS1$site_generated)
tableS1$site_generated = sub("\\.([0-9]+)", "", tableS1$site_generated)

tableS1  = tableS1  %>%
  mutate(site_generated = case_when(
    startsWith(site_generated, "500m") ~ "OC2",
    startsWith(site_generated, "3000m") ~ "OC3",
    startsWith(site_generated, "3500m") ~ "OC4",
    startsWith(site_generated, "4000m") ~ "OC5",
    startsWith(site_generated, "4500m") ~ "OC6")
  )

tableS1$site_generated = paste(tableS1$site_generated, tableS1$depth_generated, sep = ",")
tableS1$depth_generated = NULL

#specify if genome was published in alex's paper
jaffe2024 <- c("R500m500m.91", "R3000m2000m.276", "R3000m3000m.18", "R4000m1000m.46", "R3000m150m.313", "R3500m3000m.36", "R3000m3000m.264", "R4500m4000m.318", "R4000m3000m.74", "R4500m2000m.74")

# Add the previously_reported column
tableS1 <- tableS1 %>%
  mutate(previously_reported = ifelse(genome %in% jaffe2024, "Jaffe et al., 2024", "no"))

coverM_simple = select(coverM, c("Sample", "genome", "Mean"))

coverM_wide = pivot_wider(coverM_simple, names_from = Sample, values_from = Mean)

tableS1 = merge(tableS1, coverM_wide, by = "genome")



write.csv(tableS1, paste(rootdir, "refined_analysis/refined_tables/refined_tableS1.csv", sep = ""), row.names = FALSE)
```

Table S5: genome, KO, KO description, larger metabolic function, site, depth
```{r}
#read in updated auto_MAGs after manual refinement of catabolic genes using trees

auto_MAGs = read_csv(paste(rootdir, "refined_analysis/refined_auto_MAGs_edited.csv", sep = ""))

tableS5 = auto_MAGs
tableS5 = filter(tableS5, tableS5$present >= 1)
tableS5 = select(tableS5, "genome", "KO", "gene_description", "module")
tableS5 = rename(tableS5, KO_description = gene_description)
tableS5 = rename(tableS5, metabolic_function = module)
tableS5 = left_join(tableS5, gtdb_info, by = "genome")

write.csv(tableS5, paste(rootdir, "refined_analysis/refined_tables/refined_tableS5.csv", sep = ""), row.names = FALSE)

```
